\documentclass[letterpaper,12pt]{article}
\usepackage{amsmath,amssymb,amstext,amsthm}
\usepackage[margin=1in,letterpaper]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}

% Theorems numbered within sections
\newtheorem{theorem}{Theorem}[section]
% Lemmas share numbering with theorems
\newtheorem{lemma}[theorem]{Lemma}
% Propositions share numbering with theorems
\newtheorem{proposition}[theorem]{Proposition}
% Definitions numbered within sections
\newtheorem{definition}{Definition}[section]


\title{An introduction to lattice trapdoor}
\author{Ganyu Xu (g66xu)}
\date{Fall, 2023}

\begin{document}
\maketitle

\section{Introduction}
Trapdoor functions are the foundations of public-key cryptography. The most famous and popular trapdoor functions include the RSA trapdoor (it is easy to compute $m^e$ from $m$ and $e$ but difficult to recover $m$ from $m^e$ and $e$) and discrete log (it is easy to compute $g^x$ from $g, x$ but hard to recover $x$ from $g^x, g$). While these two trapdoors have been proved very successful over their decades of application, they are unfortunately both based on the hardness of integer factorization. Trapdoor functions based on other classes of hard problems have been of research interest since the early days of public-key cryptography, and they have gained increased attention as a result of the discovery of efficient quantum algorithm for integer factorization that threatens to break encryption and signature schemes based on RSA and Diffie-Hellman.

This project surveys a particular class of cryptographic trapdoor functions whose construction is based on lattices and whose security is based on the hardness of lattice problems such as the shortest vector problem (SVP) and the closest vector problem (CVP).

Due to the limit of on the write-up, some basic results of linear algebra and lattice are implicitly assumed, such as the Gram-Schmidt orthogonalized basis, the unimodular relationship between basis of the same lattice. In section 2, we will discuss the hardness of hard lattice problems by discussing the currently known best algorithm for solving them. In section 3, we will introduce the Goldreich-Golwasser-Halevi trapdoor construction and its applications.

\subsection{Notations}
Troughout this survey, vectors are denoted with bolded letters $\mathbf{v}$ while matrices are denoted by capital letters $B$. Basis is sometime represented as a matrix whose base vectors are the columns, but at other times basis is represented as an ordered sequence of base vectors $B = [\mathbf{b}_1, \mathbf{b}_2, \ldots, \mathbf{b}_n]$, whichever is more convenience. Unless otherwise specified, lattice basis $B \in \mathbb{Z}^{n \times n}$ is assumed to be full-rank integer matrix.

\subsection{Credits}
Enormous thank you to Vinod Vaikuntanathan\cite{mit-cs294} and Daniele Micciancio\cite{ucsd-cse206a} for publishing the lecture notes from their respective courses, which inspired this survey and provided much of the technical details in the proofs described in section 2 and 3.

% \input{preliminaries.tex}
\input{basis-reduction.tex}
\input{ggh-trapdoor.tex}

\section{Conclusion and final comments}
In this project, we reviewed some fundamental results of lattices, analyzed the difficulty of hard lattice problems and the efficiency of their best known solutions. We then discussed a lattice trapdoor construction that takes advantage of an exponential gap between short basis and long basis in their ability to solve the closest vector problem, and applied the trapdoor to build a public key cryptosystem and a digital signature scheme.

Both the public-key cryptosystem and the signature scheme have been cryptanalyzed \cite{nguyen1999cryptanalysis}\cite{nguyen2006learning} in their original proposed form. However, the idea of using a pair of short and long basis to build trapdoors has been extended in later studies. Gentry, Peikert, and Vaikuntanathan developed this idea into the GPV framework\cite{gentry2008trapdoors} that later led to FALCON, a digital signature scheme that was submitted to NIST's post-quantum cryptography competition. This idea was also behind NTRUSign\cite{hoffstein2003ntrusign}, whose post-quantum variant was also submitted to NIST's post-quantum cryptography competition.

\bibliographystyle{plain}
\bibliography{./references.bib}

\end{document}